Modal Abstractions for Virtualizing Memory Addresses

### 2024-07-10 examples.tex
+

\end{array}
\]
\vspace{-1em}
\caption{Global Address-Space Invariant in Figure \ref{fig:peraspaceinvariant} extended with a ghost map bookkeeping identity mappings }
  \label{fig:peraspaceinvariant_with_p2v_extension}
\end{figure*}
\textsf{pte\_walkpgdir}, as a client, holds the knowledge that there exists an identity mapping for the physical entry address (\textsf{entry}) in the root page table ($\textsf{L}_{4}$):  $\mathsf{entry} \mapsto_{\textsf{id}} \textsf{\_}$ in Specification Line 3 is a partially owned token for accessing and looking up the resources in the identity map $\Xi$ to construct the \ref\textit{virtual-to-physical} pointsto relation $\textsf{entry+KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{entry \entry\_val}$ with the virtual address (\textsf{entry+KERNBASE}) obtained by offsetting the physical address (\textsf{entry}). With this knowledge on the root-page-table-entry, we start traversing the page-table tree which requires locating the address of the next table shown in Figure \ref{fig:calltoptepteinitialize}. 
Beyond a frame, the precondition before Line 13 requires the current address space invariant, and knowledge that \textsf{entry} is mapped to a random entry value in Figure \ref{fig:peraspaceinvariant_with_p2v_extension}). By owning this capability, we can also refer to the facts in the invariant that there exists an virtual address from a range set dedicated to page-table entries, as this physical-to-virtual conversion is only guaranteed for page table memory, and obtained by offsetting the physical address ($\textsf{va} = \textsf{pa} + \textsf{KERNBASE}$. 
We see that to realize identity mappings, an additional ghost map for keeping identity mappings ($\Xi$) added to the per-address-space
invariant per Figure \ref{fig:peraspaceinvariant_with_p2v_extension}. The extended invariant owns, for each memory location dedicated to page tables, some fraction of
that physical memory, for the sake of constructing the virtual-mapping, also the fractional ownership abstracting translation (i.e. ghost translation in Figure \ref{fig:peraspaceinvariant_with_p2v_extension}), and guarantees the offset is the correct calculation.
Subtly, the operation ensures the relevant table entry is readable, but the exact portion of ownership returned must be determined by inspecting the valid bit
of the value in memory --- so full ownership is returned only for unused entries.
This is a simple piece of code whose functionality is critical and whose correctness is highly non-trivial. No prior work engages with this problem.

Concretely speaking, when we focus on the specfication of Line 13 in Figure \ref{fig:calltopteinitialize} in Figure \ref{fig:lookingupidpampping}, to read the value referenced by physical address \textsf{entry}, as shown in the first specification Line in Figure \ref{fig:lookingupidpampping}, we open our the addresss-space invariant ($\mathcal{I}\textsf{ASpace}_{\textsf{id}}(\ptablestore,\Xi,m)$), and then obtain the fractional ownership of the ghost translation and physical pointsto
% \[\sumwalkabsent \ast \mathsf{entry}\mapsto_{\mathsf{p,qfrac}}\; \vale \]
\[\fracghostmaptoken{\delta}{(\texttt{entry + KERNBASE})}{\textsf{entry}}{\textsf{qfrac}} \ast \mathsf{entry}\mapsto_{\mathsf{p,qfrac}}\; \vale \]
by using the capability $\Xi \;!!\; \textsf{entry} = \textsf{Some V}$ and left the invariant as in the state $\mathcal{I}\textsf{ASpace}_{\textsf{id}}(\ptablestore,\Xi\setminus\{\mathsf{entry}\},m)$ , and construct the virtual mapping with the resources peeled-out of the identity mapping $\Xi$ 
\[ \textsf{entry+KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{entry entry\_val} \]
under the side condition (*)
\begin{equation*}
 \ulcorner \textsf{qfrac} = 1 \leftrightarrow \; \lnot\textsf{entry\_present }(\vale) \urcorner \tag{*}
\end{equation*}
assuring that looking the identity mapping for \textsf{entry} is safe under the subtle justification which equates the full ownership to the non/presence of the entry which can only be known when investigated in Line 19in Figure \ref{fig:calltopteinitialize}.
   \specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m) \ast \textsf{rbp-8} \mapsto_{\textsf{v}} \textsf{entry} \ast \texttt{r8}  \mapsto_{\textsf{r}} \textsf{entry} \textsf{\_} \ast \ulcorner \Xi !! \textsf{entry} = \textsf{Some V} \urcorner  }_{\rtv}  \\
\sqsubseteq \\
   \specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi\setminus \{ \mathsf{entry} \},m)  \ast \textsf{rbp-8} \mapsto_{\textsf{v}} \textsf{entry} \ast \texttt{r8}  \mapsto_{\textsf{r}} \textsf{entry}  \ast  \texttt{rbp-16} \mapsto_{\textsf{v}} \textsf{next} \ast \texttt{rdi}  \mapsto_{\textsf{r}} \textsf{entry} \ast \ghostmaptoken{\delta{}s}{\rtv}{\delta}  }_{\rtv} \\
\specline{ \textsf{entry+KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{entry entry\_val} \ast \ulcorner  \textsf{qfrac} = 1 \leftrightarrow \; \lnot(\textsf{entry\_present entry\_val})\urcorner}_{\rtv} \\
\mathsf{mov} \;   [ \mathsf{r8} ],\mathsf{rdi}\\
 \specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi\setminus \{ \mathsf{entry} \},m)  \ast \textsf{rbp-8} \mapsto_{\textsf{v}} \textsf{entry} \ast \texttt{r8}  \mapsto_{\textsf{r}} \textsf{entry}  \ast  \texttt{rbp-16} \mapsto_{\textsf{v}} \textsf{next} \ast \texttt{rdi}  \mapsto_{\textsf{r}} \textsf{entry} \ast \ghostmaptoken{\delta{}s}{\rtv}{\delta}  }_{\rtv} \\
\specline{ \textsf{entry+KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{entry entry\_val} \ast \ulcorner  \textsf{qfrac} = 1 \leftrightarrow \; \lnot(\textsf{entry\_present entry\_val})\urcorner}_{\rtv} \\
 \end{array}
 \]
\caption{Peel-out the resources needed for the virtual-pte pointsto relation for physical address \textsf{entry} to be accessed in Line 7 in Figure \ref{fig:calltopteinitialize}. }
  \label{fig:lookingupidpampping}
 \end{figure*}
 \subsubsection{Ensuring Presence of an Entry}
 After obtaining the identity mapping for \textsf{entry}, we are able to load the \textsf{entry\_val} into \textsf{rdi}, and check the presence bit through
 \[ \begin{array}{l}
   \mathsf{and }\; 0x1, \;rdi \\
   \mathsf{mov } \;\mathsf{rdi}, \; \mathsf{rax} \\
   \mathsf{cmp } \; 0x0, \;\mathsf{rax}
 \end{array}
 \] Lines (17-20) in Figure \ref{fig:calltopteinitialize}.
 Accessing the presence bit and checking the value allow  us to exploit the fact in the side condition (*) in such a way that we either take the allocation path in the non-presence case (Lines 19-44 in Figure \ref{fig:calltopteinitialize}) with the knowledge of full ownership ($\textsf{qfrac} = 1$) to set the entry with proper values for the bitfields, most imporantly frame referencing freshly allocated page and presence bit, or, take non-allocation path and jump to Line 45 with fractional permission for looking up the entry values.

>>>>

\end{array}
\]
\vspace{-1em}
\caption{Global Address-Space Invariant in Figure \ref{fig:peraspaceinvariant} extended with a ghost map bookkeeping identity mappings }
  \label{fig:peraspaceinvariant_with_p2v_extension}
\end{figure*}
\textsf{pte\_walkpgdir}, as a client, holds the knowledge that there exists an identity mapping for the physical entry address (\textsf{entry}) in the root page table ($\textsf{L}_{4}$):  $\mathsf{entry} \mapsto_{\textsf{id}} \textsf{\_}$ in Specification Line 3 is a partially owned token for accessing and looking up the resources in the identity map $\Xi$ to construct the \ref\textit{virtual-to-physical} pointsto relation $\textsf{entry+KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{entry \entry\_val}$ with the virtual address (\textsf{entry+KERNBASE}) obtained by offsetting the physical address (\textsf{entry}). With this knowledge on the root-page-table-entry, we start traversing the page-table tree which requires locating the address of the next table shown in Figure \ref{fig:calltoptepteinitialize}. 
Beyond a frame, the precondition before Line 15 requires the current address space invariant, and knowledge that \textsf{entry} is mapped to a random entry value, subtly, the operation also, at least, requires that the relevant table entry is readable, but the exact portion of ownership returned must be determined by inspecting the valid bit
of the value in memory --- so full ownership is returned only for unused entries.
This is a simple piece of code whose functionality is critical and whose correctness is highly non-trivial. No prior work engages with this problem.

Concretely speaking, when we focus on the specfication of Line 15 in Figure \ref{fig:calltopteinitialize} in Figure \ref{fig:lookingupidpampping}, to read the value referenced by physical address \textsf{entry}, we need the guarantee
under the side condition (*)
\begin{equation*}
 \ulcorner \textsf{qfrac} = 1 \leftrightarrow \; \lnot\textsf{entry\_present }(\vale) \urcorner \tag{*}
\end{equation*}
assuring that looking the identity mapping for \textsf{entry} is safe under the subtle justification which equates the full ownership to the non/presence of the entry which can only be known when investigated in Line 21 in Figure \ref{fig:calltopteinitialize}.
\begin{comment}
 \begin{figure*}
   \footnotesize
   \[
 \begin{array}{l}
   \specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m) \ast \textsf{rbp-8} \mapsto_{\textsf{v}} \textsf{entry} \ast \texttt{r8}  \mapsto_{\textsf{r}} \textsf{entry} \textsf{\_} \ast \ulcorner \Xi !! \textsf{entry} = \textsf{Some V} \urcorner  }_{\rtv}  \\
\sqsubseteq \\
   \specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi\setminus \{ \mathsf{entry} \},m)  \ast \textsf{rbp-8} \mapsto_{\textsf{v}} \textsf{entry} \ast \texttt{r8}  \mapsto_{\textsf{r}} \textsf{entry}  \ast  \texttt{rbp-16} \mapsto_{\textsf{v}} \textsf{next} \ast \texttt{rdi}  \mapsto_{\textsf{r}} \textsf{entry} \ast \ghostmaptoken{\delta{}s}{\rtv}{\delta}  }_{\rtv} \\
\specline{ \textsf{entry+KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{entry entry\_val} \ast \ulcorner  \textsf{qfrac} = 1 \leftrightarrow \; \lnot(\textsf{entry\_present entry\_val})\urcorner}_{\rtv} \\
\mathsf{mov} \;   [ \mathsf{r8} ],\mathsf{rdi}\\
 \specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi\setminus \{ \mathsf{entry} \},m)  \ast \textsf{rbp-8} \mapsto_{\textsf{v}} \textsf{entry} \ast \texttt{r8}  \mapsto_{\textsf{r}} \textsf{entry}  \ast  \texttt{rbp-16} \mapsto_{\textsf{v}} \textsf{next} \ast \texttt{rdi}  \mapsto_{\textsf{r}} \textsf{entry} \ast \ghostmaptoken{\delta{}s}{\rtv}{\delta}  }_{\rtv} \\
\specline{ \textsf{entry+KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{entry entry\_val} \ast \ulcorner  \textsf{qfrac} = 1 \leftrightarrow \; \lnot(\textsf{entry\_present entry\_val})\urcorner}_{\rtv} \\
 \end{array}
 \]
 \vspace{-1em}
\caption{Peel-out the resources needed for the virtual-pte pointsto relation for physical address \textsf{entry} to be accessed in Line 7 in Figure \ref{fig:calltopteinitialize}. }
  \label{fig:lookingupidpampping}
 \end{figure*}
\end{comment}
 \subsubsection{Ensuring Presence of an Entry}
 After obtaining the identity mapping for \textsf{entry}, we are able to load the \textsf{entry\_val} into \textsf{rdi}, and check the presence bit through
 \[ \begin{array}{l}
   \mathsf{and }\; 0x1, \;rdi \\
   \mathsf{mov } \;\mathsf{rdi}, \; \mathsf{rax} \\
   \mathsf{cmp } \; 0x0, \;\mathsf{rax}
 \end{array}
 \] Lines (17-20) in Figure \ref{fig:calltopteinitialize}.
 Accessing the presence bit and checking the value allow  us to exploit the fact in the side condition (*) in such a way that we either take the allocation path in the non-presence case (Lines 19-44 in Figure \ref{fig:calltopteinitialize}) with the knowledge of full ownership ($\textsf{qfrac} = 1$) to set the entry with proper values for the bitfields, most imporantly frame referencing freshly allocated page and presence bit, or, take non-allocation path and jump to Line 45 with fractional permission for looking up the entry values.
